Parameter-less Population Pyramid

### 2014-01-17 Parameterless Population Pyramid/main.tex
Major wordsmithing to get down to 8 pages.

\maketitle
\begin{abstract}
%Maximum 200 Words
  Real world applications of evolutionary techniques are often
  hindered by the need to determine problem specific parameter
  settings.  While some previous methods have reduced or removed the
  need for parameter tuning, many do so by trading efficiency for
  general applicability.  The Parameter-less Population Pyramid (P3)
  is an evolutionary technique that both requires \textbf{no}
  parameters and is broadly effective. P3 strikes a balance between
  continuous integration of diversity with exploitative elitist
  operators, allowing it to solve easy problems quickly and hard
  problems eventually.  When compared with three optimally tuned,
  state of the art optimization techniques, P3 always finds the
  optimum at least a constant factor \emph{\textbf{faster}} across
  four benchmarks (Deceptive Trap, Deceptive Step Trap,
  Hierarchical If and only If, Rastrigin).  More importantly, on three
  randomized benchmarks (NK Landscapes, Ising Spin Glasses, MAX-SAT),
  P3 has a lower order of computational complexity as measured by
  evaluations.
  We also provide outlines for expected runtime analysis of P3,
  setting the stage for future theory based conclusions.
  Based on over 1 trillion evaluations, our results
  suggest P3 has wide applicability to a broad class of
  problems.
\end{abstract}

% A category with the (minimum) three required fields
%\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
\category{TODO}{TODO}{TODO}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Algorithms}

@@

\section{Introduction}
Providing a good solution in reasonable time to a broad class of real
world black box problems is a fundamental goal of evolutionary
computation.  Achieving this goal often requires an optimizer to make
assumptions about the nature of the problem it is solving as no
optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.

Often built into the design of a genetic algorithm (GA) are
evolutionary parameters which allow the user to tune the optimizer's
assumptions to match a specific problem.  These parameters can allow
a GA to more efficiently search the problem's landscape.  However, to
do so the user must first determine how to correctly \textbf{tune} the
parameters, typically a complex and time consuming process unto
itself.

@@

Many parameters have ``reasonable'' settings that are
unlikely to effect the order of complexity of a genetic algorithm,
with population size being a notable exception. For example, an overly small
population can cause exponential search times, while arbitrarily large
population sizes can lead to wasted evaluations and stalled search.
More importantly, setting the population size correctly relies on
difficult to discern features of each problem. 
Goldberg~\cite{goldberg:1991:gasize} showed that small
population sizes were not capable of finding good solutions, though
some practitioners~\cite{haupt:2000:optimum} have found that small
populations with a large mutation rate are effective on some practical
problems.

These complications have encouraged work to remove the need for
operator defined evolutionary parameters. In
an early example, a crossover based GA was devised which used
concurrent racing populations to remove all user specified
parameters~\cite{harik:1999:parameterlessga}.  This method of
population racing has also been paired with more advanced evolutionary
system, such as Hierarchical Bayesian
Optimization~\cite{pelikan:2004:parameterlesshboa}. The claim was that
while not optimal, this method ensured the GA would take no more than
a logarithmic factor increase in evaluations beyond optimal.  Furthermore, doing so would
ensure that these optimizers could be generally applicable without
problem specific tuning.  Experimental evidence supports this theory,
with some evidence that in general the bound is closer to a constant
factor worse than optimal.

In this work we shall introduce the Parameter-less Population Pyramid
(P3).  P3 is an iterative method for constructing a \textbf{collection} of
populations that requires no user parameters.  It leverages entropy based
linkage detection to learn how to efficiently mix solutions without
problem specific information beyond the evaluation function.  Perhaps
most importantly, unlike other parameter-less techniques it
experimentally appears to be at least a constant factor
\emph{\textbf{improvement}} over comparable optimally configured
optimization methods.

\section{The P3 Algorithm}
\label{sec-p3}
Before going into detail about how each aspect of P3 works, let us
consider an example of P3 operation. Unlike many EC techniques, P3
does not have a single population of solutions.  Instead P3 maintains
a pyramid like structure, such that each level of the pyramid has a
population of solutions, with each level representing something akin
to different generations of evolution. Consider the example evolution
presented in Figure~\ref{fig-p3}.  At the start of optimization P3
has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
is brought to a local optima using a hill climber
(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
to the population pyramid (Section~\ref{sec-pyramid}) at population
level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to continue,
so solution \#3 is randomly generated and optimized via hill climbing
to create solution \#4, which is likewise added to the population
pyrimad at population level $P_0$.  With a population of two,
crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
and \#4 creating solution \#5, which is an improvement. As \#5 used
information from $P_0$ it is added to the next level of the pyramid at
population level $P_1$.  Because of the crossover technique that P3
uses, there is no more information to be gained from the present
population. Thus, \#6 is randomly
generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
in this example because it happens to be identical to a solution
already in the pyramid.  Even though \#7 is not unique, it still
crosses with $P_0$, which does make a unique improvement \#8 that is
added to $P_1$ and mixed with that level to create \#9.  As before,
the process creates \#10, \#11, and \#12 using random generation, hill
climbing, and crossover, respectively.  In this example \#13 is not
added to $P_2$ because it is not a strict fitness improvement over
\#12.  There is not enough information in $P_2$ to improve \#13
further, so the process would begin yet again with a random solution.

@@

For our purposes, we use a First Improvement Hill Climber
(FIHC)\footnote{We use the name FHIC for an algorithm that likely has
  various names in the literature, but use it here with a descriptive
  name for clarity.}, a method of exploiting local information through
a series of modify-evaluate-accept iterations starting from a randomly
initialized solution and ending with an optimum, either local or global.
FIHC uses the following algorithm.  Consider all gene loci
$x_i$ in a random order.  For a given $x_i$ try all possible
alternative values for the gene $x_i$, evaluating solution quality
with each change.  If any alternative value is a strict fitness improvement
over the current value for $x_i$, replace $x_i$ with the improved
value.  Iteration of the process continues until no such change results in
a strict fitness improvement.

@@

FIHC solves all functions of unitation in $O(N)$ time.  This is in contrast to
methods like Steepest Ascent Hill Climber~\cite{bosman:2011:lsbbo, goldman:2012:ltga}
which require $O(N^2)$ as changes are only accepted after all neighbor solutions
are evaluated.  In general, FIHC is designed to find optima using a minimum number
of evaluations.  Furthermore, because FIHC tests loci in a random order it does
not cause any positional bias to the search.

@@

During cluster creation, P3 attempts to create a binary tree of clusters such that the leaves
represent individual gene loci and each internal node represent the subset of loci created
by joining the subsets encoded in its children.  These clusters represent linked
loci that have been detected in the genome, and are used when performing crossover.
Using the table of observed frequencies described in Section~\ref{sec-pyramid}, Equation~\ref{eq-distance}
efficiently estimates the linkage between clusters of loci using the average pairwise
entropy, as given in Equation~\ref{eq-entropy}.
\begin{equation}
  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j} 
  2 - \frac{H(C_i) + H(C_j)}{H(C_i \cup C_j)}
  \label{eq-distance}
\end{equation}
\begin{equation}
  H(C) = -\sum_{s\in S} p_c(s)\log(p_c(s))
  \label{eq-entropy}
\end{equation}

Starting from the leaves, the binary tree can be constructed using an
agglomerative method.  Each
iteration determines the pairwise linkage between each cluster of loci
using Equation~\ref{eq-distance}, and merges the two closest into a new
cluster.  The process ends when there is only a single cluster
containing all loci.  As our clustering algorithm is both convex and
satisfies the swapping lemma~\cite{gronau:2007:upgma}, this algorithm
requires $O(N^2)$ total time to correctly construct the linkage tree
from the frequency table.

Not all of  the resulting clusters in the linkage  tree are useful for
crossover.  The most obvious useless  cluster is one that contains all
the loci.   A crossover that preserves  all loci in  a solution cannot
create  a  new  individual.   We  further  remove  clusters  that  are
completely  contained   within  some  other  cluster,   such  that  if
$D(C_i,C_j)  =  0$, $C_i$  and  $C_j$  are  not used  when  performing
crossover.   $D(C_i,C_j)$ only  returns  $0$ when  the  loci in  $C_i$
perfectly predict the gene values in $C_j$ and vice versa. From this we can
conclude  that  the  settings  for  those two  loci  should  never  be
disrupted.  When  used on particularly small populations  this has the
potential  to  greatly  reduce  the  number of  clusters  used  during
crossover.  For  instance, if  $|P_i| = 1$,  no clusters exist. If
$|P_i| =  2$ at most two clusters  exist: one for all  genes where the
solutions agree and one for where they disagree.

Given a useful set of clusters, we can now define how they are used.
P3 uses crossover to recombine known information in a level of the
pyramid with a candidate solution.  The algorithm is as follows. For
each cluster select a random donor from the population.  Copy the
donor's gene values for all loci in the cluster into the candidate
solution.  Evaluate the candidate.  If the fitness of the candidate
has been reduced, revert the candidate's gene values.  It is important
to note that candidate changes are kept in the case when the fitness
is \emph{unaffected}.  This allows crossover to drift, potentially
moving through plateaus.

@@

Finally, the use of levels allows P3 to perform solution filtering
without throwing away information.  Once found, a local optima is
stored indefinitely, preserving both its linkage and genes for future
crossovers.  By splitting solutions into levels, the linkage
information and the donated genes come from similar solution
qualities.  This prevents noise in gene linkage which only exists in
low quality solutions from effecting cluster formation for high
quality solutions.  It also reduces the chance for high quality
solutions to simply overwrite new, unoptimized solutions, preserving
search diversity.

  A major advantage to the design of P3 is that it makes very weak
  assumptions about the search landscape.  In many evolutionary
  systems the crossover operator assumes that genes linked in the
  search landscape are collocated in the genome.  This introduces an
  initialization problem, changing the difficulty of the problem due
  to a, potentially unknown, ordering required on the genome.  All of
  P3's operations ignore gene ordering, removing that level of
  configuration from the user.

\subsection{Contrasted With The Generational Model}
A major advantage of P3's structure is that it avoids the need to set
a fixed population size.  In a generational model, the diversity
contained in the initial population represents a bottleneck for future
generations.  This is true because in most systems, especially on
deceptive problems, mutation rates are low enough as to make the
creation of large, multi-gene substructures unlikely after
initialization.  As such the generational model often results in a
race between decreasing diversity and increasing fitness. This means the
population size must be set large enough to contain sufficient initial
diversity and to maintain that diversity during search for just long
enough for all of it to be leveraged.

Premature convergence of a population leads to the generational model
often having two termination conditions: out of time or out of
diversity.  In the former, if a smaller population size had been used,
more of the diversity could have been exploited meaning a better
solution likely could have been reached in the same number of
evaluations.  In the latter, if a larger population size had been
used, more evaluations could have been performed increasing the
likelihood of improving solution quality.  In
either case, unless the population size is correctly set, the
generational model will likely result in waste.

As P3 has exploited all known information each time the top of the pyramid is reached,
no evaluations are wasted  on unused diversity.  Similarly, because P3
does not throw away previously optimized solutions while adding diversity, it avoids
waste caused by premature convergence. It is this essential balance of
exploitation to diversity that P3 focuses on.

In general, P3 only makes two assumptions: local structure and non uniform gene interdependence.
In order for P3 to be efficient there needs to be at least some correlation between a
solution's fitness and the fitness of its single bit neighborhood.  This is required for
the hill climber to function properly.  Under this assumption, each time the hill climber
improves a solution it also improves the expected fitness of the neighborhood of the solution.
Stated another way, this assumes that small changes in genotype result in small changes in phenotype.
The non uniform gene interdependence assumption is required by the crossover operator to
perform clustering.  Without this assumption, the clusters formed by crossover will
detect spurious relations. The resulting donation of genes between solutions will be
no better than random modification.  Both assumptions are likely to hold true for
real world problems and are similarly assumed by most generational models.

\subsection{Order of Complexity}
With the exception of crossover, P3 requires no more than $O(N)$ computations
per evaluation, implying it is limited by the time required to evaluate a solution.
In order to perform crossover, P3 must spend $O(N^2)$ time updating pairwise
frequencies each time a solution is added to a level of the pyramid.  Similarly,
it requires $O(N^2)$ time to rebuild the set of crossover clusters once the frequency
table has been updated.  Na\"ively, this leads to the algorithm itself requiring $O(N^2)$ time.
However, multiple evaluations can occur for each time a single individual is added
to a population.

Each time a random restart occurs, at minimum $N+1$ evaluations are performed.
This means each time a solution is added to $P_0$, at least $O(N)$ evaluations
must be performed.  This implies that the $N^2$ cost of the update to $P_0$ can
be amortized over these evaluations, making this update $O(N)$ per evaluation.  Note
that $N+1$ is a very weak lower bound, as this only occurs if the randomly generated
individual is already at a local optimum.  Furthermore, only restarts that result
in a unique individual cause updates to $P_0$.

To cause an update to $P_i$, all of the crossovers discovered for
$P_{i-1}$ must first be applied.  Each causes their own evaluation.
If P3 were to na\"ively use all crossovers and perform an evaluation
each time, this would cause $O(N)$ evaluations.  Again, this would
result in $O(N)$ amortized cost per update to $P_i$.  However,
Section~\ref{sec-crossover} discussed how P3 will ignore some clusters
and will also skip an evaluation if no change to the solution was made
by a crossover.  As such we are currently unable to state conclusively
that updating $P_i$ will require $O(N)$ evaluations.  That said,
empirically it appears to be true.  Furthermore, it seems counterintuitive
that intentionally wasting evaluations would be considered beneficial
to runtime.

@@

As P3 utilizes a hill climber to perform optimization, it is necessary
to show P3's added complexity is able to outperform an optimizer that
uses a hill climber alone.  Therefore as the first comparison we
define a complete optimization method based on the First Improvement
Hill Climber defined in Section~\ref{sec-hillclimber}.  This
definition is extended to match P3's style of restarting. Whenever the
hill climber obtains a solution that cannot be improved using a single
bit flip, a new solution is randomly generated and optimized.

@@

Since LTGA requires a population size parameter, we thought it most fair to
compare against LTGA with an optimized population size, even though P3
receives no such tuning. To set the population size parameter, we use
the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
size required to meet a specified success criteria.  As P3 will run until the global optima
is found or the evaluation limit is reached, we want to ensure LTGA has a similar requirement
on its success.  Therefore, using the Rule of Three~\cite{jovanovic:1997:ruleofthree}, we
can provide a bound on the probability of failure $\frac{3}{k+1}$ with $95\%$ confidence,
where $k$ is a number of runs without a failure.  Here we state that a population size
is successful if it performs 100 runs without failing to find the global optimum, meaning
the probability of that population failing to solve problems of the same class is bounded above by
$3\%$.  Note that these 100 runs are considered training, and the results from the bisection
runs are not included during the comparative testing.  Furthermore, when testing on problem
classes, bisection is performed on a different set of instances than testing.

@@

There have been other model based optimization techniques that use little to
no parameters.  Specifically LO-LIMD~\cite{posik:2011:parameterless} uses a
population size of one and a parameter free algorithm to perform optimization.
It relies on the existence of completely separable subproblems and noise free evaluation in order to determine
if two genes are linked through perturbation of their values.  While effective on
this class of problems, most real world problems do not fall into this category.
%In general, LTGA is the most powerful general optimizer with few parameters.

\subsection{\bm{$(1+(\lambda,\lambda))$}}
As a final comparison we chose the $(1+(\lambda,\lambda))$ algorithm~\cite{doerr:2013:lambdalambda}
which is currently the best, theory supported, crossover method.  Unlike P3 and LTGA,
$(1+(\lambda,\lambda))$ does not use a population of solutions or any model building
to determine problem variable linkage.  Instead it uses mutation to produce $\lambda$
offspring, selecting the best to recombine with the original parent to produce $\lambda$
more offspring using uniform crossover.  In the original work, a method for controlling
the value of $\lambda$ is provided based on offspring success, meaning there are no
parameters to be set in this algorithm.

@@

We also made minor modifications to selection and to prevent wasting evaluations.
If the best offspring produced in a generation is a mutant offspring, select that
over any of those produced using crossover.  If there is a fitness tie between the
best offspring in a generation, select the one with the maximum hamming distance
to the parent.  This encourages drift over plateaus.  While the original paper
discusses a ``mod'' version of the $\lambda$ control strategy for how to handle
when offspring of equal fitness are produced, we found this conflicted with our
method of restarting.  As such we use the original control strategy, where $\lambda$
is increased if the best offspring in a generation is not strictly better than its parent.
If an offspring produced by crossover has identical genes to either of its parents,
it is not evaluated.

\section{Test Problems}
All four of our optimization techniques share some interesting properties for
problem equivalence.  First, all are fitness scale invariant, in that changing
the fitness values without changing the relation between solution fitnesses does
not effect any of their search methods.  This is true because none of the methods
uses the fitness values directly, only when comparing the relative fitness between
two solutions using boolean operators.  This does not hold true for some optimization
methods, for instance fitness proportional selection.

>>>>

\maketitle
\begin{abstract}
%Maximum 200 Words
  Real world applications of evolutionary techniques are often
  hindered by the need to determine problem specific parameter
  settings.  While some previous methods have reduced or removed the
  need for parameter tuning, many do so by trading efficiency for
  general applicability.  The Parameter-less Population Pyramid (P3)
  is an evolutionary technique that both requires \textbf{no parameters}
  and is broadly effective. P3 strikes a balance between
  continuous integration of diversity with exploitative elitist
  operators, allowing it to solve easy problems quickly and hard
  problems eventually.  When compared with three optimally tuned,
  state of the art optimization techniques, P3 always finds the
  optimum at least a constant factor \emph{\textbf{faster}} across
  four benchmarks (Deceptive Trap, Deceptive Step Trap,
  HIFF, Rastrigin).  More importantly, on three
  randomized benchmarks (NK Landscapes, Ising Spin Glasses, MAX-SAT),
  P3 has a lower order of computational complexity as measured by
  evaluations.
  We also provide outlines for expected runtime analysis of P3,
  setting the stage for future theory based conclusions.
  Based on over 1 trillion evaluations, our results
  suggest P3 has wide applicability to a broad class of
  problems.
\end{abstract}

% A category with the (minimum) three required fields
%\category{Computing Methodologies}{Artificial Intelligence}{Search Methodologies}
\category{I.2.8}{Artificial Intelligence}{Problem Solving, Control Methods, and Search}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Algorithms, Performance, Experimentation}

@@

\section{Introduction}
Providing a good solution in reasonable time to a broad class of real
world black box problems is a fundamental goal of evolutionary
computation.  Achieving this goal requires an optimizer to make
assumptions about the nature of the problem it is solving as no
optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.
Many genetic algorithms (GAs) introduce evolutionary parameters which provide
user control of search assumptions.  These parameters can have enormous impact
on search efficiency~\cite{goldberg:1991:gasize} and can have non-obvious optimal
settings which require time consuming tuning to discover~\cite{grefenstette:1986:optimalga}.
When applied to black box problems with non-trivial evaluation functions, the
tuning process can be intractable, forcing users to rely on stock values or attempt
a small number of test values.

\begin{comment}
Often built into the design of a genetic algorithm (GA) are
evolutionary parameters which allow the user to tune the optimizer's
assumptions to match a specific problem.  These parameters can allow
a GA to more efficiently search the problem's landscape.  However, to
do so the user must first determine how to correctly \textbf{tune} the
parameters, typically a complex and time consuming process unto
itself.

@@

Many parameters have ``reasonable'' settings that are
unlikely to effect the order of complexity of a genetic algorithm,
with population size being a notable exception. For example, an overly small
population can cause exponential search times, while arbitrarily large
population sizes can lead to wasted evaluations and stalled search.
More importantly, setting the population size correctly relies on
difficult to discern features of each problem. 
Goldberg~\cite{goldberg:1991:gasize} showed that small
population sizes were not capable of finding good solutions, though
some practitioners~\cite{haupt:2000:optimum} have found that small
populations with a large mutation rate are effective on some practical
problems.
\end{comment}

The problem of parameter tuning has encouraged work to remove parameters from
user control in an effort to improve out-of-the-box GA quality. In
an early example, a crossover based GA was devised which used
concurrent racing populations which removed all user specified
parameters~\cite{harik:1999:parameterlessga}.
Population racing has also been paired with more advanced evolutionary
system, such as Hierarchical Bayesian
Optimization~\cite{pelikan:2004:parameterlesshboa}.
Without the need to set a population size, racing ensures the GA will take no more than
a logarithmic factor increase in evaluations beyond optimal.
Some experimental evidence suggests the bound is closer to a constant
factor worse than optimal.

In this work we shall introduce the Parameter-less Population Pyramid
(P3).  P3 is an iterative method for constructing a collection of
populations that requires no user parameters.  It leverages entropy based
linkage detection to learn how to efficiently mix solutions without
problem specific information beyond the evaluation function.  Perhaps
most importantly, unlike other parameter-less techniques it
experimentally appears to be at least a constant factor
\emph{\textbf{improvement}} over comparable, optimally configured,
optimization methods.

\section{The P3 Algorithm}
\label{sec-p3}
Unlike many EC techniques, P3
does not have a single population of solutions.  Instead P3 maintains
a pyramid like structure, such that each level of the pyramid has a
population of solutions, with each level representing something akin
to different generations of evolution. Consider the example evolution
presented in Figure~\ref{fig-p3}.  At the start of optimization P3
has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
is brought to a local optima using a hill climber
(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
to the population pyramid (Section~\ref{sec-pyramid}) at population
level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to continue,
so solution \#3 is randomly generated and optimized via hill climbing
to create solution \#4, which is likewise added to the population
pyramid at population level $P_0$.
Crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
and \#4 creating solution \#5. As \#5 used
information from $P_0$ it is added to the next level of the pyramid at
population level $P_1$.  Because of the crossover technique that P3
uses, there is no more information to be gained from the present
population. Thus, \#6 is randomly
generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
in this example because it happens to be identical to a solution
already in the pyramid.  Even though \#7 is not unique, it still
crosses with $P_0$, which does make a unique improvement \#8 that is
added to $P_1$ and mixed with that level to create \#9.  As before,
the process creates \#10, \#11, and \#12 using random generation, hill
climbing, and crossover, respectively.  In this example \#13 is not
added to $P_2$ because it is not a strict fitness improvement over
\#12.  There is not enough information in $P_2$ to improve \#13
further, so the process would begin yet again with a random solution.

@@

We use the First Improvement Hill Climber
(FIHC),\footnote{We use the name FIHC for descriptive clarity.  This algorithm
  likely already exists in the literature under various alternative names.} which acts as follows.
%a method of exploiting local information through
%a series of modify-evaluate-accept iterations starting from a randomly
%initialized solution and ending with an optimum, either local or global.
%FIHC uses the following algorithm.
Consider all gene loci
$x_i$ in a random order.  Evaluate the solution using all possible
values of $x_i$. If any alternative value is a strict fitness improvement
over the original value for $x_i$, replace $x_i$ with the improved
value.  Iteration of the process continues until no such change results in
a strict fitness improvement.

@@

%TODO Consider removing lsbbo citation
FIHC solves all functions of unitation in $O(N)$ time.  This is in contrast to
methods like Steepest Ascent Hill Climber~\cite{bosman:2011:lsbbo, goldman:2012:ltga}
which require $O(N^2)$ as changes are only accepted after all neighbor solutions
are evaluated.  In general, FIHC is designed to find optima using a minimum number
of evaluations.  Furthermore, because FIHC tests loci in a random order it does
not cause any positional bias to the search.

@@

During cluster creation, P3 creates a binary tree of clusters such that the leaves
represent individual gene loci and each internal node represent the subset of loci created
by joining the subsets encoded in its children.
%These clusters represent linked
%loci that have been detected in the genome, and are used when performing crossover.
Using the table of observed frequencies described in Section~\ref{sec-pyramid}, Equation~\ref{eq-distance}
calculates the linkage ``distance'' between clusters of loci using the average pairwise
entropy, as given in Equation~\ref{eq-entropy}.
\begin{equation}
  D(C_i,C_j) = \frac{1}{\left | C_i \right |\cdot \left |C_j \right|}\sum_{c_i \in C_i}\sum_{c_j \in C_j} 
  2 - \frac{H(C_i) + H(C_j)}{H(C_i \cup C_j)}
  \label{eq-distance}
\end{equation}
\begin{equation}
  H(C) = -\sum_{s\in S} p_c(s)\log(p_c(s))
  \label{eq-entropy}
\end{equation}

Starting from the leaves, the binary tree is constructed agglomeratively. Each
iteration the clusters $C_i$ and $C_j$ which minimize $D(C_i, C_j)$ are
merged to form a new cluster $C_i \cup C_j$.
The process ends when there is only a single cluster
containing all loci.  This clustering algorithm
requires $O(N^2)$ operations to construct the linkage tree
from the frequency table~\cite{gronau:2007:upgma}.

Not all of  the resulting clusters in the linkage  tree are useful for
crossover.  The most obvious useless  cluster is one that contains all
the loci.   A crossover that preserves  all loci in  a solution cannot
create  a  new  individual.   We  further  remove  clusters  that  are
completely  contained   within their direct superset,   such  that  if
$D(C_i,C_j) = 0$, $C_i$  and  $C_j$ are  not used  when  performing
crossover.   $D(C_i,C_j)$ only  returns  $0$ when  the  loci in  $C_i$
perfectly predict the gene values in $C_j$ and vice versa. From this we can
conclude  that  the relative settings  for  those  loci  should  never  be
disrupted.  When  used on particularly small populations  this has the
potential  to  greatly  reduce  the  number of  clusters  used  during
crossover.  For  instance, if  $|P_i| = 1$,  no clusters exist. If
$|P_i| =  2$ at most two clusters  exist: one for all  genes where the
solutions agree and one for where they disagree.
This means that Figure~\ref{fig-p3} is infeasible as crossover
cannot make \#5 and \#9 unique from \{\#2, \#4\} and \{\#5, \#8\}, respectively.

P3 uses crossover to recombine known information in a level of the
pyramid with a candidate solution.  For
each cluster select a random donor from the population.  Copy the
donor's gene values for all loci in that cluster into the candidate
solution.  Evaluate the candidate.  If the fitness of the candidate
has been reduced, revert the candidate's gene values.  It is important
to note that candidate changes are kept in the case when the fitness
is \emph{unaffected}.  This allows crossover to drift, potentially
moving through fitness plateaus.

@@

Finally, the use of levels allows P3 to perform solution filtering
without throwing away information.  Once found, a local optima is
stored indefinitely, preserving both its linkage and genes for future
crossovers.  By splitting solutions into levels, the linkage
information and the donated genes come from similar solution
qualities.  This prevents noise in gene linkage which only exists in
low quality solutions from effecting cluster formation for high
quality solutions.  It also reduces the chance that optimized
solutions will overwrite newer solutions, preserving
search diversity.

P3 makes very weak
assumptions about the search landscape.  In many evolutionary
systems the crossover operator assumes linked genes
are collocated in the genome.  This requires the user
to configure genome structure based on gene linkage information they
are unlikely to have. All of
P3's operations ignore gene ordering, removing that level of
configuration from the user.

In general, P3 makes two assumptions about the search landscape:
local structure and non uniform gene interdependence.
For the hill climber to be effective, there should be at least some
correlation between a solution's fitness and the fitness of its single
bit neighborhood. Stated another way, this assumes that small changes
in genotype result in small changes in phenotype. The non uniform gene
interdependence assumption is required by the crossover operator to
perform clustering.  Without this assumption, the clusters formed by
crossover will detect spurious relations. The resulting donation of genes
between solutions will be no better than random modification.  Both
assumptions are likely to hold true for real world problems and are similarly
assumed by most evolutionary optimizers.

\subsection{Contrasted With The Generational Model}
A major advantage of P3's structure is that it avoids the need to set
a fixed population size.  In a generational model, the diversity
contained in the initial population represents a bottleneck for future
generations.  This is true because in most systems, especially on
deceptive problems, mutation rates are too low
to create large, multi-gene substructures after
initialization.  As such the generational model often results in a
race between decreasing diversity and increasing fitness. 
%This means the
%population size must be set large enough to contain sufficient initial
%diversity and to maintain that diversity during search for just long
%enough for all of it to be leveraged.

Premature convergence of a population causes the generational model
to have two termination conditions: out of time or out of
diversity.  In the former, if a smaller population size had been used,
less evaluations could have been spent supporting diversity allowing more
focus on exploitation.
In the latter, if a larger population size had been
used, more evaluations could have been performed increasing the
likelihood of improving solution quality.  In
either case, unless the population size is correctly set, the
generational model will likely result in waste.

In contrast, P3 has exploited all known information each time the top of the pyramid is reached;
no evaluations are wasted  on unused diversity.  Similarly, because P3
does not throw away previously optimized solutions while adding diversity, it avoids
waste caused by premature convergence.

\subsection{Order of Complexity Per Evaluation}
With the exception of crossover, P3 requires no more than $O(N)$ computations
per evaluation, implying it is limited by the time required to evaluate a solution.
In order to perform crossover, P3 must spend $O(N^2)$ operations updating pairwise
frequencies each time a solution is added to a level of the pyramid.  Similarly,
it requires $O(N^2)$ time to rebuild the set of crossover clusters once the frequency
table has been updated.  Na\"ively, this leads to the algorithm itself requiring $O(N^2)$ time.
However, multiple evaluations can occur for each time a single individual is added
to a population.

Each time a random restart occurs, at minimum $N+1$ evaluations are performed.
This means each time a solution is added to $P_0$, at least $O(N)$ evaluations
must be performed.  This implies that the $O(N^2)$ cost of the update to $P_0$ can
be amortized over these evaluations, making this update $O(N)$ per evaluation.
%Note
%that $N+1$ is a very weak lower bound, as this only occurs if the randomly generated
%individual is already at a local optimum.  Furthermore, only restarts that result
%in a unique individual cause updates to $P_0$.

To cause an update to $P_i$, all of the crossovers discovered for
$P_{i-1}$ must first be applied, each causing their own evaluation.
If P3 were to na\"ively use all crossovers and perform an evaluation
each time, this would cause $O(N)$ evaluations.  Again, this would
result in $O(N)$ amortized cost to update $P_i$.  However,
Section~\ref{sec-crossover} discussed how P3 will ignore some clusters
and will also skip an evaluation if no change to the solution was made
by a cluster.  As such we are currently unable to state conclusively
that updating $P_i$ will require $O(N)$ operations per evaluation.  That said,
empirically it appears to be true.  Furthermore, it seems counterintuitive
that intentionally wasting evaluations would be considered beneficial
to runtime.

@@

As P3 utilizes a hill climber to perform optimization, it is necessary
to show P3's added complexity is able to outperform an optimizer that
uses a hill climber alone.  Therefore as the first comparison we
define a complete optimization method based on the First Improvement
Hill Climber from Section~\ref{sec-hillclimber}.  This
definition is extended to match P3's style of restarting. Whenever the
hill climber obtains a solution that cannot be improved using a single
bit flip, a new solution is randomly generated and optimized.

@@

To ensure maximal LTGA quality, we use the bisection method~\cite{goldman:2012:ltga}
to minimally set LTGA's population size.  To be considered successful, a
population size must result in 100 runs without premature convergence.
According to the Rule of Three~\cite{jovanovic:1997:ruleofthree}, this bounds
LTGA's failure rate to less than $\frac{3}{100+1} \approx 3\%$ with 95\% confidence.
Note that runs used in setting the population are considered testing and were not
used during analysis.  On randomized problems, training was performed on a disjoint
set of problem instances.

\begin{comment}
Since LTGA requires a population size parameter, we thought it most fair to
compare against LTGA with an optimized population size, even though P3
receives no such tuning. To set the population size parameter, we use
the bisection method~\cite{goldman:2012:ltga}, which determines the minimum population
size required to meet a specified success criteria.  As P3 will run until the global optima
is found or the evaluation limit is reached, we want to ensure LTGA has a similar requirement
on its success.  Therefore, using the Rule of Three~\cite{jovanovic:1997:ruleofthree}, we
can provide a bound on the probability of failure $\frac{3}{k+1}$ with $95\%$ confidence,
where $k$ is a number of runs without a failure.  Here we state that a population size
is successful if it performs 100 runs without failing to find the global optimum, meaning
the probability of that population failing to solve problems of the same class is bounded above by
$3\%$.  Note that these 100 runs are considered training, and the results from the bisection
runs are not included during the comparative testing.  Furthermore, when testing on problem
classes, bisection is performed on a different set of instances than testing.
\end{comment}

@@

%TODO Consider removing this
There have been other model based optimization techniques that use little to
no parameters.  Specifically LO-LIMD~\cite{posik:2011:parameterless} uses a
population size of one and a parameter free algorithm to perform optimization.
It relies on the existence of completely separable subproblems and noise free evaluation in order to determine
if two genes are linked through perturbation of their values.  While effective on
this class of problems, most real world problems do not fall into this category.
%In general, LTGA is the most powerful general optimizer with few parameters.

\subsection{\bm{$(1+(\lambda,\lambda))$}}
As a final comparison we chose the $(1+(\lambda,\lambda))$ algorithm~\cite{doerr:2013:lambdalambda}
which is currently the best theory based crossover method.  Unlike P3 and LTGA,
$(1+(\lambda,\lambda))$ does not use a population of solutions or any model building
to determine problem variable linkage.  Instead it uses mutation to produce $\lambda$
offspring, selecting the best to recombine with the original parent to produce $\lambda$
more offspring using uniform crossover.  In the original work, a method for controlling
the value of $\lambda$ is provided based on offspring success, meaning there are no
parameters to be set in this algorithm.

@@

We also made minor modifications to selection and to prevent wasting evaluations.
If the best offspring produced in a generation is a mutant offspring, select that
over any of those produced using crossover.  If there is a fitness tie between the
best offspring in a generation, select the one with the maximum hamming distance
to the parent.  This encourages drift over plateaus.  While the original paper
discusses a ``mod'' version of the $\lambda$ control strategy for how to handle
when offspring of equal fitness are produced, we found this conflicted with our
method of restarting.  As such we use the original control strategy. %, where $\lambda$
%is increased if the best offspring in a generation is not strictly better than its parent.
If an offspring produced by crossover has identical genes to either of its parents,
it is not evaluated.
